---
title: "Principle Component Aanalysis"
output: html_document
---
## Data
The data you will be using comes from the Assistments online intelligent tutoring system (https://www.assistments.org/). It describes students working through online math problems. Each student has the following data associated with them:

- id
- prior_prob_count: How many problems a student has answered in the system prior to this session
- prior_percent_correct: The percentage of problems a student has answered correctly prior to this session
- problems_attempted: The number of problems the student has attempted in the current session
- mean_correct: The average number of correct answers a student made on their first attempt at problems in the current session
- mean_hint: The average number of hints a student asked for in the current session
- mean_attempt: The average number of attempts a student took to answer a problem in the current session
- mean_confidence: The average confidence each student has in their ability to answer the problems in the current session

## Start by uploading the data
```{r setup}

library(dplyr)
library(tidyr)
library(ggplot2)
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)

D1 <- read.csv("Assistments-confidence.csv")

#We won't need the id variable, so remove that.
D1 <- select(D1, -c(id))

```

## Create a correlation matrix of the relationships between the variables, including correlation coefficients for each pair of variables/features.

```{r correlation matrix}
#You can install the corrplot package to plot some pretty correlation matrices (sometimes called correlograms)
#install.library("corrplot")
library(corrplot)

#Generate pairwise correlations
COR <- cor(D1)

corrplot(COR, order="AOE", method="circle", tl.pos="lt", type="upper",        
tl.col="black", tl.cex=0.6, tl.srt=45, 
        addCoef.col="black", addCoefasPercent = TRUE,
        sig.level=0.50, insig = "blank")
ggsave("Correlation Plot.jpg", path = file.path(getwd(), "Images"))

#Study your correlogram image and save it, you will need it later
```

## Create a new data frame with the mean_correct variable removed

```{r Remove mean_correct variables}
D2 <- select(D1, -mean_correct)

#The, scale and center your data for easier interpretation
D2 <- scale(D2, center = TRUE, scale = TRUE)

```

## Now run the PCA on the new data frame

```{r PCA analysis}

pca <- prcomp(D2, scale = TRUE)
pca

#note that the squared vector of each principal component sums to 1: 
sum((pca$rotation^2)[,1])
sum((pca$rotation^2)[,6])

#thus, the squared eigenvector is the proportion of variance that each item contributes to the respective component

```

## Although princomp does not generate the eigenvalues directly for us, we can print a list of the standard deviation of the variance accounted for by each component.

```{r Calculating Eigenvalue}
pca$sdev

#To convert this into variance accounted for we can square it, these numbers are proportional to the eigenvalue

pca$sdev^2

#A summary of our pca will give us the proportion of variance accounted for by each component

summary(pca) #how to calculate the porportion of variance of each component?

#We can look at this to get an idea of which components we should keep and which we should drop
#scree plot
plot(pca, type = "lines")
```

## Think about which components you would drop and make a decision
#looking at the plot, I would retain 2 factors since that is the point where there is a large bed (or change in slope). Looking at the summary statistics, we can see that the including 2 components would capture .46 percent of the varaition in the items.

## Part II

```{r}
#Now, create a data frame of the transformed data from your pca.

D3 <- as.data.frame(pca$x)

#Attach the variable "mean_correct" from your original data frame to D3.

D4 <- cbind(D3, as.data.frame(D1$mean_correct))

#Now re-run your scatterplots and correlations between the transformed data and mean_correct. If you had dropped some components would you have lost important infomation about mean_correct?

COR2 <- cor(D4)
corrplot(COR2, order="AOE", method="circle", tl.pos="lt", type="upper",        
tl.col="black", tl.cex=0.6, tl.srt=45, 
        addCoef.col="black", addCoefasPercent = TRUE,
        sig.level=0.50, insig = "blank")

#Retaining two factors works well as can be seen in the correlation matrix. The big problem is that componenet 6 has a lot of information to contribute to mean correct, so I would retain 6 componenets since that the 6th compnent has the most information in it abotu mean correct
  
```
## Now print out the eigenvectors (often called loadings) for the components you generated:

```{r}
pca$rotation

#Examine the eigenvectors, notice that they are a little difficult to interpret. It is much easier to make sense of them if we make them proportional within each component

loadings <- abs(pca$rotation) #abs() will make all eigenvectors positive

sweep(loadings, 2, colSums(loadings), "/") #sweep() computes each row as a proportion of the column. (There must be a way to do this with dplyr()?)

#Now examine your components and try to come up with substantive descriptions of what some might represent?
#Component 1: This has to deal with issues related to grasping the problem, as hints and attempts are clustering together
#Component 2: This has to deal with getting the problems correct
#Component 3: this has to deal with the single factor of confidence, which is to some degree related to getting prior problems crrect
#I would not interpret the later components as these first three explain the main chunks of variation

#You can generate a biplot to help you, though these can be a bit confusing. They plot the transformed data by the first two components. Therefore, the axes represent the direction of maximum variance. Then mapped onto this point cloud are the original directions of the variables, depicted as red arrows. It is supposed to provide a visualization of which variables "go together". Variables that possibly represent the same underlying construct point in the same direction.  

biplot(pca)

#Calculate values for each student that represent these your composite variables and then create a new correlogram showing their relationship to mean_correct.

PC_scores <- pca$x

D5 <- cbind(PC_scores, as.data.frame(D1$mean_correct))
COR3 <- cor(D5)

corrplot(COR3, order="AOE", method="circle", tl.pos="lt", type="upper",        
tl.col="black", tl.cex=0.6, tl.srt=45, 
        addCoef.col="black", addCoefasPercent = TRUE,
        sig.level=0.50, insig = "blank")
```
# Part III  
## Also in this repository is a data set and codebook from Rod Martin, Patricia Puhlik-Doris, Gwen Larsen, Jeanette Gray, Kelly Weir at the University of Western Ontario about people's sense of humor. Can you perform a PCA on this data?

```{r PCA on Western Ontario dataset}
dat <- read.csv("humor_data.csv")

#This dataset is crazy, so doing a PCA will reduce it into compnent parts and make it manageable 

#step 1 - put all variables on same scale by stand
dat2 <- scale(dat, center = TRUE, scale = TRUE)

#step 2 - run PCA
pca <- prcomp(dat2, scale = TRUE)
pca

#Step 3 - use scree plot to figure out optimal number of components
plot(pca, type = "lines")
sum <- summary(pca)
sum$importance

#according to the summary and scree plot, choosing component 2 or 6 are the points of inflections. 
#I would retain 6 components - we are reducing a dataset of 39 variables down to only 6 componenets are are retaining over 50% of the total variation in the sytem.

```

